```{r}
library(nlsr)
pK1 <- -1; pK2 <- 3; pK3 <- 7.55 # literature values
HEPTOT <- 0.05
kw <- 1e-14
source('definitions.R')
```

Differentiating the synthetic $\mathrm{pH}$ values will let us know how
every parameter impacts the value of the function. Thankfully, `nlsr`
already has the necessary tools for that, accessible from plain **R**:

```{r}
SID <- seq(-5e-2,4e-2,length=1000)
e <- quote(pHm(SID, HEPTOT, pK1, pK2, pK3))
pH <- numericDerivR(e, paste0('pK', 1:3), central = TRUE)
plot(SID, pH, type = 'l')
```

One minor detail: the values are different enough in their orders of
magnitude to warrant the use of logarithmic scale. There will be very
small negative numbers which we'll replace with `NA` to avoid warnings.

```{r}
g <- attr(pH, 'gradient')
g[g <= 0 & abs(g) < 1e-9] <- NA
par(mar = c(5, 6, 4, 2) + .1)
matplot(SID, g, type = 'l', lty = 1, col = 2:4, log = 'y', ylab = quote(
	frac(partialdiff*'pH', partialdiff*pK[j])
))
legend('bottomright', lty = 1, col = 2:4, legend = sapply(1:3,
	function(j) bquote(pK[.(j)])
))
```

This shows very clearly that the contribution of $\mathrm{p}K_1$ is very
minor and is only noticeable at significantly negative $\mathrm{SID}$,
which imply low $\mathrm{pH}$ values, which cannot be reliably measured.
